---
title: "Taxi Trip Duration Prediction"
author: "Ravinder Ram"
date: "25/12/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Files
```{r}
getwd()
list.files()
setwd()
```

### Import dataset
```{r}
library(tidyverse)
train <- read_csv("train.csv")
test <- read_csv("test.csv")
str(train)
str(test)
```
### Solution -1 
make a ml model that predict datetime but there is no ml algo do this type of task so i escape it

### Solution -2
calculate distance between two location then divide avg speed to get trip duration

function which measure distance between two location
```{r}
# make a function
earth.dist <- function (pickup_long, pickup_lat, dropoff_long, dropoff_lat) {
rad <- pi/180
a1 <- pickup_lat * rad
a2 <- pickup_long * rad
b1 <- dropoff_lat * rad
b2 <- dropoff_long * rad
dlon <- b2 - a2
dlat <- b1 - a1
a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
c <- 2 * atan2(sqrt(a), sqrt(1 - a))
R <- 6378.145
d <- R * c
return(d)
}
```


test the function
```{r}
earth.dist(-73.98813, 40.73203, -73.99017, 40.75668)
```

apply this function to test dataset
add new col loc_dist on test dataset
```{r}
test <- test %>% mutate(loc_dist = earth.dist(test$pickup_longitude,                                             test$pickup_latitude,
                            test$dropoff_longitude, test$dropoff_latitude))
```

add new col trip_duration on test dataset 
```{r}
test <- test %>% mutate(trip_duration = (loc_dist * 3600)/12.5)
```

now check your position in kaggle leaderboard
### submit in kaggle
```{r}
submission <- data.frame(id = test$id, trip_duration = test$trip_duration )

#write csv
write.csv(submission, "kaggle.csv", row.names = F)
```

### Solution-3

add new col loc_dist on test dataset
```{r}
test <- test %>% mutate(loc_dist = earth.dist(test$pickup_longitude,                                             test$pickup_latitude,
                            test$dropoff_longitude, test$dropoff_latitude))
```

add new col by calculate earth distance for train data
```{r}
train <- train %>% mutate(loc_dist = earth.dist(train$pickup_longitude,                                             train$pickup_latitude,
                            train$dropoff_longitude, train$dropoff_latitude))
```

check structure and new col which you created recently
```{r}
glimpse(train)
```

extract some information from datetime column
```{r}
library(lubridate)
# pickup_datetime
# extracts week
train$p_week <- train$pickup_datetime %>% wday()
test$p_week <- test$pickup_datetime %>% wday()
# weekend - saturday, sunday office holiday
# new col weekend(5) & no_weekend(0)
train$weekend <- if_else(train$p_week > 1 &
                           train$p_week < 7, 0, 5)
test$weekend <- if_else(test$p_week > 1 &
                           test$p_week < 7, 0, 5)

# extract months
train$p_month <- train$pickup_datetime %>% month()
test$p_month <- test$pickup_datetime %>% month()

# extract hours
train$greeting <- train$pickup_datetime %>% hour()
test$greeting <- test$pickup_datetime %>% hour()
# rush hour - 7 to 9a.m  and 6 to 8p.m
# rush - 65  no_rush - 45
train$rush <- if_else(train$greeting >= 7 &
                        train$greeting <= 9, 65, 45)
test$rush <- if_else(test$greeting >= 7 &
                        test$greeting <= 9, 65, 45)

train$rush <- if_else(train$greeting >= 18 &
                        train$greeting <= 20, 65, 45)
test$rush <- if_else(test$greeting >= 18 &
                        test$greeting <= 20, 65, 45)
# extract minute
train$minute <- train$pickup_datetime %>% minute()
test$minute <- test$pickup_datetime %>% minute()

## early morning(4 to 8a.m) - 25
train$greeting <- if_else(train$greeting >= 4 & train$greeting <= 8, 25L, train$greeting)
test$greeting <- if_else(test$greeting >= 4 & test$greeting <= 8, 25L, test$greeting)
## morning(8a.m to 12) - 26
train$greeting <- if_else(train$greeting >= 9 & train$greeting <= 12, 26L, train$greeting)
test$greeting <- if_else(test$greeting >= 9 & test$greeting <= 12, 26L, test$greeting)
## afternoon(12 to 17) - 27
train$greeting <- if_else(train$greeting >= 13 & train$greeting <= 17, 27L, train$greeting)
test$greeting <- if_else(test$greeting >= 13 & test$greeting <= 17, 27L, test$greeting)
## evening(17 to 20) - 28
train$greeting <- if_else(train$greeting >= 18 & train$greeting <= 20, 28L, train$greeting)
test$greeting <- if_else(test$greeting >= 18 & test$greeting <= 20, 28L, test$greeting)
## night(20 to 23) - 29
train$greeting <- if_else(train$greeting >= 21 & train$greeting <= 23, 29L, train$greeting)
test$greeting <- if_else(test$greeting >= 21 & test$greeting <= 23, 29L, test$greeting)

## late night
train$greeting <- if_else(train$greeting >= 0 & train$greeting <= 3, 30L, train$greeting)
test$greeting <- if_else(test$greeting >= 0 & test$greeting <= 3, 30L, test$greeting)

as.factor(train$greeting)
as.factor(test$greeting)

## you can also try for dropoff time in train data but dropoff time column 
## not availbale in test data so i escape it
```
### data explore and wrangle

see relation between passenger_count and trip_duration
```{r}
train %>% ggplot(aes(x = as.factor(passenger_count), y = trip_duration)) + 
  geom_boxplot()
```

make new col van(90) and taxi(9)
```{r}
train <- train %>% mutate(car = if_else(passenger_count > 4, 90, 9 ))
test <- test %>% mutate(car = if_else(passenger_count > 4, 90, 9 ))
```


calculate manhattan distance
```{r}
# make a function
earth.dist <- function (pickup_long, pickup_lat, dropoff_long, dropoff_lat) {
rad <- pi/180
a1 <- pickup_lat * rad
a2 <- pickup_long * rad
b1 <- dropoff_lat * rad
b2 <- dropoff_long * rad
dlon <- b2 - a2
dlat <- b1 - a1
a <- (sin(dlat/2))^2 + cos(a1) * cos(b1) * (sin(dlon/2))^2
c <- 2 * atan2(sqrt(a), sqrt(1 - a))
R <- 6378.145
d <- R * c
return(d)
}

test <- test %>% mutate(loc_dist = earth.dist(test$pickup_longitude,                                             test$pickup_latitude,
                            test$dropoff_longitude, test$dropoff_latitude))

train <- train %>% mutate(loc_dist = earth.dist(train$pickup_longitude,                                             train$pickup_latitude,
                            train$dropoff_longitude, train$dropoff_latitude))
```





### build model
using h2o automl

first remove unncessary column from train and test data
```{r}
train <- train %>% select(-c(pickup_datetime, pickup_longitude, pickup_latitude,
                             dropoff_datetime, dropoff_latitude, dropoff_longitude, ))

test <- test %>% select(-c(pickup_datetime, pickup_longitude, pickup_latitude,
                             dropoff_latitude, dropoff_longitude))
```


we are going to make model so convert your all character column into factor
and any colum which you think it should be as factor for ex:- vendor_id

```{r}
# for train data
train$vendor_id <- as.factor(train$vendor_id)
train$passenger_count <- as.factor(train$passenger_count)
train$store_and_fwd_flag <- as.factor(train$store_and_fwd_flag)
train$weekend <- as.factor(train$weekend)
train$p_week <- as.factor(train$p_week)
train$p_month <- as.factor(train$p_month)
train$greeting <- as.factor(train$greeting)
train$rush <- as.factor(train$rush)
train$minute <- as.factor(train$minute)
train$car <- as.factor(train$car)

# for test data
test$vendor_id <- as.factor(test$vendor_id)
test$passenger_count <- as.factor(test$passenger_count)
test$store_and_fwd_flag <- as.factor(test$store_and_fwd_flag)
test$weekend <- as.factor(test$weekend)
test$p_week <- as.factor(test$p_week)
test$p_month <- as.factor(test$p_month)
test$greeting <- as.factor(test$greeting)
test$car <- as.factor(test$car)
test$rush <- as.factor(test$rush)
test$minute <- as.factor(test$minute)
```

define independent variable and dependent variable list
```{r}
y <- "trip_duration"
x <- setdiff(names(train), y)
```



```{r}
library(h2o)
h2o.init(max_mem_size = "2G")
```





```{r}
# convert into h2o frame
train <- as.h2o(train)
test <- as.h2o(test)
```



automl
```{r}
#automl
aml <- h2o.automl(x=x, y = y,
           training_frame = train,
           max_runtime_secs = 180)

lb <- aml@leaderboard
lb
```


predict
```{r}
#predict
y_pred <- h2o.predict(aml, test[,-1])
y_pred
```

### submit in kaggle
```{r}
submission <- data.frame(id = as.data.frame(test$id),
                         trip_duration = as.data.frame(y_pred))

#write csvas
write.csv(submission, "kaggle.csv", row.names = F)
```